Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-11 draft.tex
Added Reference to Supplementary Material at the introduction

In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques to derive insight \cite{Klemm2014VIS}.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
%%
\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
%%
\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

@@

\section*{\uppercase{Appendix}}

\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{blind.dnsalias.com}.

>>>>

In this paper, we present a thorough analysis of the image-derived data from lumbar back spine datasets and analyze their discriminating power for different characteristics of the disease.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques to derive insight \cite{Klemm2014VIS}.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis
%%
\item Detection of correlations between image-based and socio-demographic as well as medical parameters for an improved hypothesis generation
%%
\item Identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R-Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To assess a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Epidemiological Data}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies, analyzing one specific disease and its influences on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim to collect as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found under the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - identical}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

@@

%\section*{\uppercase{Appendix}}
%\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{blind.dnsalias.com}.
